Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-07-11 chapters/Paper_five.tex
Minor changes

\section{Introduction}\label{sec:intro_five}
Liquid droplets are an ubiquitous phenomenon which has raised scientific curiosity for decades.
Among sprays, sliding drops or sessile drops, there is a vast zoo of different hydrodynamic problems.
The coalescence of liquid droplets or the lack of it is one of these problems. 
Both scenarios can be observed in many instances of our everyday life and in various industrial processes.
Coalescence of sessile droplets happens, for example, when vapor condenses.
During this process, small droplets nucleate randomly and quickly assemble in larger drops, either naturally~\cite{PhysRevA.43.1906} or due to artificial factors such as surface structure or patterning~\cite{C1SM06219K}. 
This effect can effectively be used in so-called fog harvesting devices, where ambient water vapor is condensed and collected in structures which promote droplet coarsening~\cite{zhang2015inkjet, shi2018fog}.
Apart from fog harvesting, coalescence or rather the control of it plays an important part in inkjet printing and printable electronics~\cite{jo2009evaluation, singh2010inkjet, Kim_2005, Luechinger_2008}, but also in various microfluidic devices for mixing purposes at low Reynolds numbers ($Re < 1$)~\cite{https://doi.org/10.1002/pen.760352206, doi:10.1063/1.858199}.

@@

\section{Theory}\label{sec:theory_five}
The theoretical approach we use to study this system is the lubrication approximation~\cite{Reynolds, RevModPhys.69.931, PhysRevE.63.011208}.
Applying this approximation to the Navier-Stokes equation yields the thin film equation, which for a singular horizontal dimension reads~\cite{RevModPhys.81.739, RevModPhys.81.1131, THIELE2014399}
\begin{equation}\label{eq:thin_film_simple}
    \partial_t h = \partial_x \left(\frac{h^3}{3\mu}\partial_x p\right),
\end{equation}
where $\frac{h^3}{3\mu}$ is the mobility $M(h)$\footnote{$m(h) = h^3/3\mu$ is true only for a no-slip velocity boundary condition.}, $h(x,t)$ is the thickness of the film at time $t$ and position $x$, $\mu$ is the liquid's viscosity, and $p$ it's pressure.
The pressure accounts for both the surface tension and the correct fluid substrate behavior (wetting)~\cite{PhysRevE.100.033313}.
We therefore write the pressure as
\begin{equation}\label{eq:pressure_five}
    p = \gamma \partial_x^2 h + \Pi(h),
\end{equation}
where the first term is the (1D) Laplacian of the liquid-vapor interface and $\Pi(h)$ is the disjoining pressure~\cite{RevModPhys.69.931, RevModPhys.81.739, Peschka9275, PhysRevE.63.011208},
\begin{equation}\label{eq:disjoin}
    \Pi(h) = K(\gamma,\theta)\left[\left(\frac{h_{\ast}}{h}\right)^n - \left(\frac{h_{\ast}}{h}\right)^m\right].
\end{equation}
The prefactor $K(\gamma,\theta)\propto \gamma(1-\cos(\theta))$ encodes the wettability and as such the equilibrium contact angle $\theta_{\text{eq.}}$.
This prefactor can be understood as a linking factor to the Hamaker constant ($\mathcal{A}$) with~~\cite{PhysRevE.93.013120, bestehorn20033d, van1988interfacial}
\begin{equation}
    \mathcal{A} = 6\pi h_{\ast}^3 K(\theta).    
\end{equation}
Additionally, $h_{\ast}$ defines the thickness of the precursor layer, therefore the thickness where $\Pi(h_{\ast}) = 0$.
The pair of powers $(n,m)$ need to satisfy $n > m$ and $m > 1$.
For an in deep presentation of the derivation we refer to the work by Schwartz and Eley and Bonn et al.~\cite{SCHWARTZ1998173, RevModPhys.81.739}.

@@

The addition of a surface tension gradient alters the pressure balance.
For two different but miscible liquids, Riegler and Lazar associated Eq.~(\ref{eq:P_Bernulli}) with the difference in surface tension~\cite{doi:10.1021/la800630w}
\begin{equation}\label{eq:P_bernulli_riegler}
    P_{iner.} \sim \frac{\rho (\delta\gamma)^2}{\mu^2}, 
\end{equation}
where $\delta\gamma = \gamma_1 - \gamma_2$ and $\gamma_i$ are the two surface tensions of two different liquids. 
Karptischka and Riegler later formulated an elegant theoretical explanation for the flow of the non-coalescing twin droplet state~\cite{PhysRevLett.109.066103}.
In their derivation, they assume a quasistatic thickness ($\partial_t h \approx 0$) and a constant motion proportional to the capillary number 
\begin{equation}\label{eq:sys_cap_vel}
    Ca = \frac{\mu v}{\gamma},
\end{equation} 
with $v$ being a characteristic velocity that preserves the bridge's profile. 
The starting point is Eq.~(\ref{eq:thin_with_marangoni}) in the quasistatic limit, 
\begin{equation}\label{eq:pressure_noncoal}
    \partial_x\left(\frac{h^3}{3\mu}\partial_x p + \frac{h^2}{2\mu}\partial_x\gamma\right) = 0.
\end{equation}
Integrating the above equation and isolating the pressure gradient $\partial_x p = \gamma\partial_x^3 h$ yields~\cite{RevModPhys.69.931, PhysRevLett.109.066103}
\begin{equation}\label{eq:karpitschka_h3}
    \partial_x^3 h = \frac{3}{h^2}Ca - \frac{3}{2h\gamma}\partial_x\gamma.
\end{equation}
In the absence of a surface tension gradient we have $\partial_x^3 h \geq 0$.
With a surface tension gradient, however, $\partial_x^3 h$ can change sign~\cite{PhysRevLett.109.066103}.

Our model in addition to the above derivation uses a pressure given by Eq.~(\ref{eq:pressure_five}), therefore modifying Eq.~(\ref{eq:karpitschka_h3}) slightly, 
\begin{equation}\label{eq:stefan_h3}
    \partial_x^3 h \approx \frac{3}{h^2}Ca - \frac{3}{2h\gamma}\partial_x\gamma - \partial_x\Pi(h).
\end{equation}
where the derivative of the disjoining pressure appears. 
The velocity field inside the droplets (film) can be computed using $\partial_x^3 h$~\cite{RevModPhys.69.931},
\begin{equation}\label{eq:Oron_correct}
    u(z) = \frac{z}{\mu}\left[\partial_x\gamma - \left(\frac{z}{2} - h\right)\gamma\partial_x^3 h\right].
\end{equation}
We therefore assume that the derivative of the disjoining pressure has an effect on the flow, at least in the region close to the bridge.

>>>>

\section{Introduction}\label{sec:intro_five}
Liquid droplets are an ubiquitous phenomenon which has raised scientific curiosity for decades.
Among sprays, sliding drops or sessile drops, there is a vast zoo of different hydrodynamic problems.
The coalescence of liquid droplets or the lack of it is one of these problems. 
Both scenarios can be observed in many instances of our everyday life and in various industrial processes.
Coalescence of sessile droplets happens, for example, when vapor condenses.
During this process, small droplets nucleate randomly and quickly assemble in larger drops, either naturally~\cite{PhysRevA.43.1906} or due to artificial factors such as surface structure or patterning~\cite{C1SM06219K}. 
This effect can effectively be used in so-called fog harvesting devices, where ambient water vapor is condensed and collected in structures which promote droplet coarsening~\cite{zhangInkjetPrintingDirect2015, shiFogHarvestingHarps2018}.
Apart from fog harvesting, coalescence or rather the control of it plays an important part in inkjet printing and printable electronics~\cite{jo2009evaluation, singh2010inkjet, Kim_2005, Luechinger_2008}, but also in various microfluidic devices for mixing purposes at low Reynolds numbers ($Re < 1$)~\cite{https://doi.org/10.1002/pen.760352206, doi:10.1063/1.858199}.

@@

\section{Theory}\label{sec:theory_five}
The theoretical approach we use to study this system is the lubrication approximation~\cite{Reynolds, oronLongscaleEvolutionThin1997, PhysRevE.63.011208}.
Applying this approximation to the Navier-Stokes equation yields the thin film equation, which for a singular horizontal dimension reads~\cite{RevModPhys.81.739, RevModPhys.81.1131, thielePatternedDepositionMoving2014}
\begin{equation}\label{eq:thin_film_simple}
    \partial_t h = \partial_x \left(\frac{h^3}{3\mu}\partial_x p\right),
\end{equation}
where $\frac{h^3}{3\mu}$ is the mobility $M(h)$\footnote{$m(h) = h^3/3\mu$ is true only for a no-slip velocity boundary condition.}, $h(x,t)$ is the thickness of the film at time $t$ and position $x$, $\mu$ is the liquid's viscosity, and $p$ it's pressure.
The pressure accounts for both the surface tension and the correct fluid substrate behavior (wetting)~\cite{PhysRevE.100.033313}.
We therefore write the pressure as
\begin{equation}\label{eq:pressure_five}
    p = \gamma \partial_x^2 h + \Pi(h),
\end{equation}
where the first term is the (1D) Laplacian of the liquid-vapor interface and $\Pi(h)$ is the disjoining pressure~\cite{oronLongscaleEvolutionThin1997, RevModPhys.81.739, Peschka9275, PhysRevE.63.011208},
\begin{equation}\label{eq:disjoin}
    \Pi(h) = K(\gamma,\theta)\left[\left(\frac{h_{\ast}}{h}\right)^n - \left(\frac{h_{\ast}}{h}\right)^m\right].
\end{equation}
The prefactor $K(\gamma,\theta)\propto \gamma(1-\cos(\theta))$ encodes the wettability and as such the equilibrium contact angle $\theta_{\text{eq.}}$.
This prefactor can be understood as a linking factor to the Hamaker constant ($\mathcal{A}$) with~~\cite{PhysRevE.93.013120, bestehorn20033d, van1988interfacial}
\begin{equation}
    \mathcal{A} = 6\pi h_{\ast}^3 K(\theta).    
\end{equation}
Additionally, $h_{\ast}$ defines the thickness of the precursor layer, therefore the thickness where $\Pi(h_{\ast}) = 0$.
The pair of powers $(n,m)$ need to satisfy $n > m$ and $m > 1$.
For an in deep presentation of the derivation we refer to the work by Schwartz and Eley and Bonn et al.~\cite{SCHWARTZ1998173, RevModPhys.81.739}.

@@

The addition of a surface tension gradient alters the pressure balance.
For two different but miscible liquids, Riegler and Lazar associated Eq.~(\ref{eq:P_Bernulli}) with the difference in surface tension~\cite{doi:10.1021/la800630w}
\begin{equation}\label{eq:P_bernulli_riegler}
    P_{iner.} \sim \frac{\rho (\delta\gamma)^2}{\mu^2}, 
\end{equation}
where $\delta\gamma = \gamma_1 - \gamma_2$ and $\gamma_i$ are the two surface tensions of two different liquids. 
Karptischka and Riegler later formulated an elegant theoretical explanation for the flow of the non-coalescing twin droplet state~\cite{PhysRevLett.109.066103}.
In their derivation, they assume a quasistatic thickness ($\partial_t h \approx 0$) and a constant motion proportional to the capillary number 
\begin{equation}\label{eq:sys_cap_vel}
    Ca = \frac{\mu v}{\gamma},
\end{equation} 
with $v$ being a characteristic velocity that preserves the bridge's profile. 
The starting point is Eq.~(\ref{eq:thin_with_marangoni}) in the quasistatic limit, 
\begin{equation}\label{eq:pressure_noncoal}
    \partial_x\left(\frac{h^3}{3\mu}\partial_x p + \frac{h^2}{2\mu}\partial_x\gamma\right) = 0.
\end{equation}
Integrating the above equation and isolating the pressure gradient $\partial_x p = \gamma\partial_x^3 h$ yields~\cite{oronLongscaleEvolutionThin1997, PhysRevLett.109.066103}
\begin{equation}\label{eq:karpitschka_h3}
    \partial_x^3 h = \frac{3}{h^2}Ca - \frac{3}{2h\gamma}\partial_x\gamma.
\end{equation}
In the absence of a surface tension gradient we have $\partial_x^3 h \geq 0$.
With a surface tension gradient, however, $\partial_x^3 h$ can change sign~\cite{PhysRevLett.109.066103}.

Our model in addition to the above derivation uses a pressure given by Eq.~(\ref{eq:pressure_five}), therefore modifying Eq.~(\ref{eq:karpitschka_h3}) slightly, 
\begin{equation}\label{eq:stefan_h3}
    \partial_x^3 h \approx \frac{3}{h^2}Ca - \frac{3}{2h\gamma}\partial_x\gamma - \partial_x\Pi(h).
\end{equation}
where the derivative of the disjoining pressure appears. 
The velocity field inside the droplets (film) can be computed using $\partial_x^3 h$~\cite{oronLongscaleEvolutionThin1997},
\begin{equation}\label{eq:Oron_correct}
    u(z) = \frac{z}{\mu}\left[\partial_x\gamma - \left(\frac{z}{2} - h\right)\gamma\partial_x^3 h\right].
\end{equation}
We therefore assume that the derivative of the disjoining pressure has an effect on the flow, at least in the region close to the bridge.

### 2023-07-11 chapters/Paper_four.tex
Minor changes

Small amounts of liquid deposited on a substrate are an everyday phenomenon.
From a theoretical point of view this represents a modelling challenge, due to the multiple scales involved: from the molecular interactions among the three phases (solid substrate, liquid film and surrounding gas) to the hydrodynamic flows.
An efficient way to deal with this problem is via the thin-film equation:
\begin{equation}\label{eq:thin_film_Joss}
    \partial_t h = \nabla\cdot(M(h)\nabla p),
\end{equation}
where $h$ is the film thickness, $M(h)$ is a thickness dependent mobility and $p$ is the pressure inside the film.
Solving the thin film equation directly is however a difficult task as it is a fourth order degenerate PDE~\cite{becker2003complex}.
\textit{Swalbe.jl} approaches the problem from a different angle.
Instead of directly solving the thin film equation we use a novel method based on a class lattice Boltzmann models~\cite{krueger2017}, originally developed to simulate shallow water flows~\cite{Salmon:1999:0022-2402:503}.
This allows us to benefit from the simplicity of the lattice Boltzmann algorithm which makes it straightforward to parallelize the code and run it on accelerator devices.
Choosing appropriate forces it is possible to simulate complex problems.
Among them is the dewetting of a patterned substrates as shown in Fig.~\ref{fig:logo}.
It is as well possible to simulate low contact angle droplets out of equilibrium to probe relaxation experiments, e.g. the Cox-Voinov or Tanner's law~\cite{RevModPhys.81.739}.
Due to a disjoining pressure model for the three phase contact line droplets can not only relax towards their equilibrium they can slide as well~\cite{PhysRevE.100.033313}.
All of this can be coupled with thermal fluctuations to study the stochastic thin film equation~\cite{shah_van_steijn_kleijn_kreutzer_2019}.

@@

Wilczek et al. used \href{https://www.dune-project.org/}{\textbf{DUNE}} to study the dynamics of an ensemble of sliding drops in ref.~\cite{PhysRevLett.119.204501}.
DUNE is a software suite written in \textit{C++} that solves partial differential equations with a grid based approach~\cite{sander2020dune}.
Therefore DUNE is not limited to the problem of thin film flows, interested readers may visit the project's home page.

>>>>

Small amounts of liquid deposited on a substrate are an everyday phenomenon.
From a theoretical point of view this represents a modelling challenge, due to the multiple scales involved: from the molecular interactions among the three phases (solid substrate, liquid film and surrounding gas) to the hydrodynamic flows.
An efficient way to deal with this problem is via the thin-film equation:
\begin{equation}\label{eq:thin_film_Joss}
    \partial_t h = \nabla\cdot(M(h)\nabla p),
\end{equation}
where $h$ is the film thickness, $M(h)$ is a thickness dependent mobility and $p$ is the pressure inside the film.
Solving the thin film equation directly is however a difficult task as it is a fourth order degenerate PDE~\cite{becker2003complex}.
\textit{Swalbe.jl} approaches the problem from a different angle.
Instead of directly solving the thin film equation we use a novel method based on a class lattice Boltzmann models~\cite{krugerLatticeBoltzmannMethod2017}, originally developed to simulate shallow water flows~\cite{Salmon:1999:0022-2402:503}.
This allows us to benefit from the simplicity of the lattice Boltzmann algorithm which makes it straightforward to parallelize the code and run it on accelerator devices.
Choosing appropriate forces it is possible to simulate complex problems.
Among them is the dewetting of a patterned substrates as shown in Fig.~\ref{fig:logo}.
It is as well possible to simulate low contact angle droplets out of equilibrium to probe relaxation experiments, e.g. the Cox-Voinov or Tanner's law~\cite{RevModPhys.81.739}.
Due to a disjoining pressure model for the three phase contact line droplets can not only relax towards their equilibrium they can slide as well~\cite{PhysRevE.100.033313}.
All of this can be coupled with thermal fluctuations to study the stochastic thin film equation~\cite{shah_van_steijn_kleijn_kreutzer_2019}.

@@

Wilczek et al. used \href{https://www.dune-project.org/}{\textbf{DUNE}} to study the dynamics of an ensemble of sliding drops in ref.~\cite{wilczekSlidingDropsEnsemble2017}.
DUNE is a software suite written in \textit{C++} that solves partial differential equations with a grid based approach~\cite{sander2020dune}.
Therefore DUNE is not limited to the problem of thin film flows, interested readers may visit the project's home page.

### 2023-07-11 chapters/Paper_one.tex
Minor changes

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host
of natural and technological settings \cite{de2003capillarity,Focke}. Therefore,
understanding and controlling their stability and dynamics is a central problem
for fundamental physics, as well as for applied research in process engineering
and nanotechnology~\cite{RevModPhys.69.931,Utada}. Coating processes, for instance,
rely crucially on the mutual affinity of liquid and surface (i.e. on
wettability properties). When the liquid film is sufficiently thin, in fact, it
can become unstable, leading to the dewetting of the coated
area~\cite{RevModPhys.81.739}. From the modelling point of view, the challenge
consists in the fact that the physics of thin films is intrinsically
multiscale, for it involves phenomena ranging from the molecular scale 
at the three phase contact line, to the micro-/nano-metric size of the 
film thickness, to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic
approach would be, obviously, unfeasible, if hydrodynamic regimes are to be
explored. It clearly appears that some degree of model order reduction is required. Most hydrodynamic models of thin liquid films, in the framework of the
lubrication theory, simplify the complexity of the full 3D Navier-Stokes
equations \cite{Navier,Stokes}
to one scalar transport equation (the lubrication equation) for the film thickness field
$h(x,t)$~\cite{ReynoldsLubr,RevModPhys.69.931,RevModPhys.81.1131,Mitlin}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left(Q(h)\nabla p_{\mbox{\tiny{film}}}\right)
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the
boundary condition for the velocity at the surface (for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with
$\mu$ being the dynamic viscosity), and $p_{\mbox{\tiny{film}}}$ is the film
pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{becker2003complex}.
Moreover, an ever-growing number of microfluidic problems requires to cope with
complex fluids rather than simple liquids, i.e. fluids with non-trivial
internal microstructure and/or complex non-Newtonian rheological
behaviour (e.g. colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical
method for simulating thin film hydrodynamics, versatile for the inclusion of
multiphysics features, is, thus, an ongoing endeavour.

@@

literature~\cite{RevModPhys.69.931,THIELE2014399}, where here we use the expression

@@

are commonly chosen values in the literature~\cite{moulton_lega_2013,RevModPhys.69.931}

@@

PhysRevLett.119.204501, wedershoven2014infrared}). 

>>>>

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host
of natural and technological settings \cite{de2003capillarity,Focke}. Therefore,
understanding and controlling their stability and dynamics is a central problem
for fundamental physics, as well as for applied research in process engineering
and nanotechnology~\cite{oronLongscaleEvolutionThin1997,Utada}. Coating processes, for instance,
rely crucially on the mutual affinity of liquid and surface (i.e. on
wettability properties). When the liquid film is sufficiently thin, in fact, it
can become unstable, leading to the dewetting of the coated
area~\cite{RevModPhys.81.739}. From the modelling point of view, the challenge
consists in the fact that the physics of thin films is intrinsically
multiscale, for it involves phenomena ranging from the molecular scale 
at the three phase contact line, to the micro-/nano-metric size of the 
film thickness, to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic
approach would be, obviously, unfeasible, if hydrodynamic regimes are to be
explored. It clearly appears that some degree of model order reduction is required. Most hydrodynamic models of thin liquid films, in the framework of the
lubrication theory, simplify the complexity of the full 3D Navier-Stokes
equations \cite{Navier,Stokes}
to one scalar transport equation (the lubrication equation) for the film thickness field
$h(x,t)$~\cite{ReynoldsLubr,oronLongscaleEvolutionThin1997,RevModPhys.81.1131,Mitlin}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left(Q(h)\nabla p_{\mbox{\tiny{film}}}\right)
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the
boundary condition for the velocity at the surface (for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with
$\mu$ being the dynamic viscosity), and $p_{\mbox{\tiny{film}}}$ is the film
pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{becker2003complex}.
Moreover, an ever-growing number of microfluidic problems requires to cope with
complex fluids rather than simple liquids, i.e. fluids with non-trivial
internal microstructure and/or complex non-Newtonian rheological
behaviour (e.g. colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical
method for simulating thin film hydrodynamics, versatile for the inclusion of
multiphysics features, is, thus, an ongoing endeavour.

@@

literature~\cite{oronLongscaleEvolutionThin1997,thielePatternedDepositionMoving2014}, where here we use the expression

@@

are commonly chosen values in the literature~\cite{moulton_lega_2013,oronLongscaleEvolutionThin1997}

@@

wilczekSlidingDropsEnsemble2017, wedershoven2014infrared}). 

### 2023-07-11 chapters/Paper_three.tex
Minor changes

\section{Introduction} Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{gross1980fluid,szeri2010fluid,DERYCK1998278,doi:10.1146/annurev.fluid.31.1.347,DASILVASOBRINHO19991204,singh2010inkjet,jo2009evaluation,Wijshoff2010}. 
Moreover, the continuously growing technological interest for lab-on-a-chip devices~\cite{C6LC00387G,Focke} as well as for printable electronics~\cite{Kim_2005, Luechinger_2008} and printable photovoltaics~\cite{Brabec2001,RH20}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{ButtEtAl_Langmuir2018,GuoGuo_RSCAdv2016}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{LiuChem2005, XiCSR2010}, involving smart materials such as polymer brushes~\cite{CohenStuartEtAl_NatMat2010,ayres2007stimuli,YongEtAl_Mat2018}, thermal-responsive hydrogels~\cite{ChenEtAl_SM2010}, light-responsive molecules and microstructures~\cite{IchimuraEtAl_Science2000,DelormeEtAl_Langmuir2005,OscuratoEtAl_AAMI2017} or processes such as electrowetting~\cite{MugeleEtAl_JPCM2005}.
Probably the simplest, yet non-trivial, modelling of a switchable substrate can be realized by a space and time dependent wettability pattern~\cite{D0SM02082F}. 
While a consistent body of theoretical/computational work has been devoted to processes on static heterogeneous substrates, though, the time dependent case is still almost unexplored, with few relevant exceptions focusing on single droplet spreading and sliding~\cite{D0SM02082F,GrawitterStark2,ThieleHartmann} or limited to an analysis of the linear regime~\cite{suman2006dynamics}.

@@

\section{Method} In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{ReynoldsLubr,RevModPhys.69.931,RevModPhys.81.1131} 

@@

where $U_{\Theta}$ is the retraction speed $U_{\Theta} = \frac{\gamma \Theta^3}{9\mu}$~\cite{Edwardse1600183}, with $\Theta = \max_{\mathbf{x}}\theta(\mathbf{x})$.

@@

\section{Acknowledgements} SZ and JH acknowledge financial support from the Deutsche Forschungsgemeinschaft (DFG) within the priority program SPP2171 ``Dynamic Wetting of Flexible, Adaptive, and Switchable Substrates'', project HA-4382/11.
\newpage
\section{Supplemental Material}.
\subsection{Minkowski's structure metric $q_2$ in an extended parameter space.}
\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{graphics/SupMatFig_1.pdf}
    \includegraphics[width=0.4\textwidth]{graphics/SupMatFig_2.pdf}
    \caption{LEFT PANEL. Minkowski's structure metric $q_2$ for three different wavelengths $\lambda=64 h_0$, $\lambda=128 h_0$ and 
    $\lambda=256 h_0$ (as in figure 4 of the main text) and for $\Gamma=1.5$ (all other parameters are as in the main text). 
    The time interval during which $q_2 \approx 1$ signals the emergence of rivulets, also for $\lambda = 64 h_0$ which is comparable with the spinodal wavelength $\lambda_s \approx 70 h_0$. RIGHT PANEL. Comparison of the Minkowski's structure metric $q_2$ for $\delta\theta=5^{\circ}$ and $\delta\theta=10^{\circ}$, 
    $\Gamma = 15$ and $\lambda = 256 h_0$.}
    \label{fig:q2_difflambda}
\end{figure}
\noindent In this section we test the robustness of the observation of the rivulet 
state over a wider parameter space and in particular at changing: 1) the 
pattern wavelength $\lambda$ and 2) the heterogeneity amplitude $\delta \theta$.
To this aim we have run simulations with $\Gamma=1.5$, $\lambda/h_0 = 64, 128$, and with $\Gamma = 15$, $\lambda = 256 h_0$, $\delta \theta = 5^{\circ}$, respectively.
In Fig.~\ref{fig:q2_difflambda} (left panel) we report the measurements of the Minkowski's structure metric $q_2$, whose 
increase from zero up to a plateauing value of $q_2 \approx 1$ indicates the emergence of rivulets, for 
the different $\lambda$'s ($\lambda=256 h_0$, which is the value considered in the main text, is also reported 
for comparison).
Analogously, in the right panel, we report $q_2$ as a function of time for  
$\delta \theta = 5^{\circ}$  and $\delta \theta = 10^{\circ}$ (the value used in the main text). 
We see that the $q_2$ metric attains the value $q_2 \approx 1$, signalling the rivulet state, albeit over 
a time interval shorter than for $\delta \theta = 10^{\circ}$, i.e. the rivulets lifetime decreases with 
$\delta \theta$. This was somehow expected, since obviously (and also in the static case) 
the patterning looses effectiveness as the contact angle mismatch is reduced (see, e.g.~\cite{PhysRevLett.84.931}).

>>>>

\section{Introduction} Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{gross1980fluid,szeri2010fluid,DERYCK1998278,doi:10.1146/annurev.fluid.31.1.347,dasilvasobrinhoStudyDefectsUltrathin1999,singh2010inkjet,jo2009evaluation,Wijshoff2010}. 
Moreover, the continuously growing technological interest for lab-on-a-chip devices~\cite{C6LC00387G,Focke} as well as for printable electronics~\cite{Kim_2005, Luechinger_2008} and printable photovoltaics~\cite{Brabec2001,RH20}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{ButtEtAl_Langmuir2018,GuoGuo_RSCAdv2016}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{LiuChem2005, XiCSR2010}, involving smart materials such as polymer brushes~\cite{CohenStuartEtAl_NatMat2010,ayres2007stimuli,YongEtAl_Mat2018}, thermal-responsive hydrogels~\cite{ChenEtAl_SM2010}, light-responsive molecules and microstructures~\cite{IchimuraEtAl_Science2000,DelormeEtAl_Langmuir2005,OscuratoEtAl_AAMI2017} or processes such as electrowetting~\cite{MugeleEtAl_JPCM2005}.
Probably the simplest, yet non-trivial, modelling of a switchable substrate can be realized by a space and time dependent wettability pattern~\cite{D0SM02082F}. 
While a consistent body of theoretical/computational work has been devoted to processes on static heterogeneous substrates, though, the time dependent case is still almost unexplored, with few relevant exceptions focusing on single droplet spreading and sliding~\cite{D0SM02082F,GrawitterStark2,ThieleHartmann} or limited to an analysis of the linear regime~\cite{suman2006dynamics}.

@@

\section{Method} In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{ReynoldsLubr,oronLongscaleEvolutionThin1997,RevModPhys.81.1131} 

@@

where $U_{\Theta}$ is the retraction speed $U_{\Theta} = \frac{\gamma \Theta^3}{9\mu}$~\cite{edwardsNotSpreadingReverse2016}, with $\Theta = \max_{\mathbf{x}}\theta(\mathbf{x})$.

@@

\section{Acknowledgements} SZ and JH acknowledge financial support from the Deutsche Forschungsgemeinschaft (DFG) within the priority program SPP2171 ``Dynamic Wetting of Flexible, Adaptive, and Switchable Substrates'', project HA-4382/11.
\newpage
\section{Supplemental Material}.
\subsection{Minkowski's structure metric \texorpdfstring{$q_2$}{hmm} in an extended parameter space.}
\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{graphics/SupMatFig_1.pdf}
    \includegraphics[width=0.4\textwidth]{graphics/SupMatFig_2.pdf}
    \caption{LEFT PANEL. Minkowski's structure metric $q_2$ for three different wavelengths $\lambda=64 h_0$, $\lambda=128 h_0$ and 
    $\lambda=256 h_0$ (as in figure 4 of the main text) and for $\Gamma=1.5$ (all other parameters are as in the main text). 
    The time interval during which $q_2 \approx 1$ signals the emergence of rivulets, also for $\lambda = 64 h_0$ which is comparable with the spinodal wavelength $\lambda_s \approx 70 h_0$. RIGHT PANEL. Comparison of the Minkowski's structure metric $q_2$ for $\delta\theta=5^{\circ}$ and $\delta\theta=10^{\circ}$, 
    $\Gamma = 15$ and $\lambda = 256 h_0$.}
    \label{fig:q2_difflambda}
\end{figure}
\noindent In this section we test the robustness of the observation of the rivulet 
state over a wider parameter space and in particular at changing: 1) the 
pattern wavelength $\lambda$ and 2) the heterogeneity amplitude $\delta \theta$.
To this aim we have run simulations with $\Gamma=1.5$, $\lambda/h_0 = 64, 128$, and with $\Gamma = 15$, $\lambda = 256 h_0$, $\delta \theta = 5^{\circ}$, respectively.
In Fig.~\ref{fig:q2_difflambda} (left panel) we report the measurements of the Minkowski's structure metric $q_2$, whose 
increase from zero up to a plateauing value of $q_2 \approx 1$ indicates the emergence of rivulets, for 
the different $\lambda$'s ($\lambda=256 h_0$, which is the value considered in the main text, is also reported 
for comparison).
Analogously, in the right panel, we report $q_2$ as a function of time for  
$\delta \theta = 5^{\circ}$  and $\delta \theta = 10^{\circ}$ (the value used in the main text). 
We see that the $q_2$ metric attains the value $q_2 \approx 1$, signalling the rivulet state, albeit over 
a time interval shorter than for $\delta \theta = 10^{\circ}$, i.e. the rivulets lifetime decreases with 
$\delta \theta$. This was somehow expected, since obviously (and also in the static case) 
the patterning looses effectiveness as the contact angle mismatch is reduced (see, e.g.~\cite{PhysRevLett.84.931}).
